# Interactive component

We first added the college names to the corresponding ADM2019 rows.
```{r}
hd2019 <- read.csv("./Data/hd2019.csv")
adm2019 <- read.csv("./Data/adm2019.csv")
#nrow(adm2019)
newhd2019 <- filter(hd2019, UNITID %in% adm2019$UNITID)
#nrow(newhd2019)
newhd2019 <- newhd2019[order(newhd2019$UNITID),]
adm2019 <- adm2019 [order(adm2019 $UNITID),]
adm2019$INSTNM <- newhd2019$INSTNM
write.csv( adm2019, "./Data/newadm2019.csv", row.names=FALSE)
```

In the below interactive plot, you can select a data point around your SAT/ACT score. Hovering over nearby points will display the college name, and few of the admission requirements for the college. The legend to understand the requirements is below the plot. 

(please ignore some mess around the plot, we are still debugging that)
```{r}
library(r2d3)
```

```{d3 }
<iframe src="./docs/edavd3.html" width="560" height="600"></iframe>
```


1: Required;
2: Recommended;
3: Neither required nor recommended;
5: Considered but not required;
